Ultrafast Control of Spin Interactions in Honeycomb Antiferromagnetic Insulators

### 2019-05-02 Paper/paper.tex
paper update

% !TEX spellcheck = en-US
\documentclass[aps,prl,twocolumn,amsmath,amssymb,nobibnotes]{revtex4-1}%
\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{bm}
\usepackage{color}
\usepackage{amsmath}
\usepackage{epstopdf}
%\usepackage{caption}
%\usepackage{subcaption}
\usepackage{fancybox}
\usepackage{amsfonts}
\usepackage{amssymb}%
\usepackage{natbib}
\usepackage[toc,page]{appendix}
\usepackage[symbol]{footmisc}
%\usepackage[style=numeric-comp]{biblatex}
\setcounter{MaxMatrixCols}{30}
%\usepackage{subcaption}

@@

\begin{document}
\title{Ultrafast Control of Spin-Spin Interactions in Two-Dimensional Magnets}

\author{Juan M. Losada}
\affiliation{Center for Quantum Spintronics, Department of Physics, Norwegian University of Science and Technology, NO-7491 Trondheim, Norway}
\author{Arne Brataas}
\affiliation{Center for Quantum Spintronics, Department of Physics, Norwegian University of Science and Technology, NO-7491 Trondheim, Norway}
\author{Alireza Qaiumzadeh}
%\thanks{Corresponding author: alireza.qaiumzadeh@ntnu.no}
\affiliation{Center for Quantum Spintronics, Department of Physics, Norwegian University of Science and Technology, NO-7491 Trondheim, Norway}
%\affiliation{Institute for Advanced Studies in Basic Science (IASBS), Zanjan, Iran}

\begin{abstract}
Light provides ultrafast control of exchange and Dzyaloshinskii-Moriya interactions, we consider two-dimensional honeycomb lattices described by the Kane-Mele-Hubbard model at half-filling and in the strongly correlated limit, i.e., the Mott insulator regime of a canted antiferromagnet. Based on the Floquet theory, we demonstrate that changing the amplitude and frequency of circularly polarized laser pulses, tunes the amplitude, the sign and even the ratio between these two spin interactions. Furthermore, the renormalizations of the spin interactions are helicity independent. Our results pave the way for ultrafast optical manipulation of spins in recently discovered two-dimensional magnetic materials.
\end{abstract}

@@

\textit{Introduction}.--- The discovery of all-optical control of the order parameters in antiferromagnetic (AFM) and ferromagnetic (FM) materials by ultrashort, intense and high-frequency laser pulses has transformed spintronics to a new era of ultrafast magnetism \cite{kimel2005ultrafast,Rasing}. Although the underlying microscopic mechanism remains unclear, there have been many attempts to uncover its origins. Recently, it has been theoretically proposed that laser pulses can modify and even reverse the sign of the exchange interaction \cite{PhysRevLett.113.057201,PhysRevLett.115.075301,Mentink2015,PhysRevLett.118.097701,Stepanov2017,Kitamura,MentinkReview,Barbeau2018}, the strongest spin interaction in magnetic systems. This finding has been supported by recent experiments \cite{2018Natur.553..481G,Mikhaylovskiy2015,Perakis,PhysRevLett.93.197403}. Another important spin interaction is the Dzyaloshinskii-Moriya interaction (DMI), an antisymmetric exchange interaction in inversion symmetry broken systems, where the chiral symmetry of the system is broken \cite{dzyaloshinsky1958, moriya1960,Alireza-DMI1, Alireza-DMI2}. While this interaction is much weaker than the exchange interaction, its presence has important consequences in magnetic materials and causes weak ferromagnetism in AFM materials \cite{dzyaloshinsky1958, moriya1960}, topological objects like chiral skyrmions \cite{2009Sci...323..915M,2017NatRM...217031F,AlirezaSk1, AlirezaSk2} and chiral domain walls \cite{Thiaville_2012,Parkin,AlirezaSW}, and exotic phase of topological magnon insulators \cite{Owerre2016, Elyasi2018, Chen2018, Owerre1, Owerre2}. The ratio between these two interactions controls the tilting angle of the canted spins. Finding a mechanism to tune this ratio enables new phenomena in ultrafast spin dynamics and switching \cite{Mikhaylovskiy2015,Mentink2015,Stepanov2017,Kitamura}.

Another recent important breakthrough in spintronics is the discovery of two-dimensional (2D) van der Waals AFM and FM materials with metallic, semiconducting and insulating band structures \cite{2D1,2D2}. The advantage of low dimensionality and magnetic order in the same material enables novel spintronics devices with exceptional performances.

In this Letter, we show that intense and high-frequency laser pulses can change the spin-spin interactions and the ratio between the exchange interaction and the DMI dramatically in a wide class of 2D magnetic materials described by the Kane-Mele-Hubbard model.

\textit{Model Hamiltonian}.---
Electrons in a 2D planar honeycomb lattice in the presence of electron-electron interactions can be described by the Kane-Mele-Hubbard model \cite{Kane2005,PhysRevLett.107.010401,PhysRevLett.106.100403,PhysRevB.82.075106,PhysRevB.84.235149,Griset2012,Auerbach}.
The Hamiltonian in the absence of external perturbations is the sum of the kinetic term $\hat{\mathcal{H}}_t$, the intrinsic SOI $\hat{\mathcal{H}}_{\text{SOI}}$ and the Coulomb interaction between the electrons modelled by the extended Hubbard interaction $\hat{\mathcal{H}}_{\text{int}}$,
\begin{equation}
\label{MKMH}
\hat{H}_0 = \hat{\mathcal{H}}_{\text{K}} + \hat{\mathcal{H}}_{\text{SOI}} + \hat{\mathcal{H}}_{\text{int}},
\end{equation}
where
\begin{align}
&\hat{\mathcal{H}}_{\text{K}} = - t_{1}\sum_{\langle i,j \rangle, \tau} \hat{c}_{i \tau}^\dagger \hat{c}_{j \tau} - t_{2}\sum_{\langle \langle i,j \rangle \rangle, \tau} \hat{c}_{i \tau}^\dagger \hat{c}_{j \tau}, \label{Ht} \\
&\hat{\mathcal{H}}_{\text{SOI}}  = i \Delta \sum_{\langle \langle i,j \rangle \rangle, \tau,\tau'} \nu_{ij}\sigma^z_{\tau, \tau'}\hat{c}_{i \tau}^\dagger \hat{c}_{j \tau'} \label{Hsoi}, \\
&\hat{\mathcal{H}}_{\text{int}} = U_{00}\sum_{i=1} \hat{n}_{i\uparrow}\hat{n}_{i\downarrow} + \frac{1}{2}\sum_{\langle i,j \rangle, \tau \tau'} V_{ij}\hat{n}_{i\tau}\hat{n}_{j\tau'} \label{Hint}.
\end{align}
Here $\langle\cdot\rangle$ and $\langle\langle\cdot\rangle\rangle$ denote, respectively, nearest neighbors (NN) and next-nearest neighbours (NNN), $\hat{c}_{i \sigma}^\dagger$ ($ \hat{c}_{i \sigma}$) are fermionic creation (annihilation) operators for an electron on site $i$ and spin state $\tau=\{\uparrow,\downarrow\}$, $t_1$ and $t_2$ are NN and NNN hopping amplitudes, $\Delta$ is the intrinsic SOI parameter, $\nu_{ij}=\pm 1$ depending on whether the electron traversing from $i$ to $j$ makes a right ($+1$) or a left turn ($-1$), $\sigma^{z}$ is the $z$-component of Pauli matrices $\bm{\sigma}$, $U_{00}$ and $V_{ij}$ are the on-site and NN Coulomb interactions.
The presence of the intrinsic NNN SOI, Eq. (\ref{Hsoi}), reduces the $\mathrm{SU}(2)$ symmetry of the original Hubbard model to the $\mathrm{U}(1)$ spin group. In buckled noncoplanar honeycomb lattices and systems with structural inversion asymmetry, the presence of, respectively, NNN and NN Rashba SOIs, further reduces the symmetry to $Z_2$.

\begin{figure}[t]
\centering
\includegraphics[width=0.6\columnwidth]{kmh.png}
\caption{A honeycomb cell with NN hopping $t_1$, NNN hopping $t_2$, intrinsic SOI $\Delta$, and $\nu_{ij} = \pm 1$ for clockwise and counterclockwise hoping.}
\label{fig1}
\vspace*{-6pt}
\end{figure}

Using the variational principle, it has been shown that the NN Coulomb interaction can be approximated by a renormalized local interaction $U =U_{00} - \bar{V}$, where $\bar{V}$ is a weighted average of the NN Coulomb interaction \cite{Schuler2013,Stepanov2017}. Thus, we only take into account the local Coulomb interaction in our total Hamiltonian and express the interaction part of the Hamiltonian in Eq. (\ref{Hint}) as
\begin{equation}
\hat{\mathcal{H}}_{\text{int}} \approx U\hat{d},
\end{equation}
where we define the doublon number operator $\hat{d} = \sum_{i=1} \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}$ with eigenvalues $d$. We denote the projection operator to the subspace of states with eigenvalue $d$ (i.e. states with exactly $d$ double occupancies) by $\hat{P}_d$. We are interested in the strongly correlated regime $U\gg t_{1(2)}$. In such a strong-coupling limit any state with a nonzero number of double occupancies ($d \neq 0$) has a much larger energy than those with no double occupancies ($d=0$). We obtain an effective Hamiltonian acting on the $d=0$ subspace by standard second-order perturbation theory in the hopping terms. Using the following relations,
\begin{align}
&\hat{c}_{i \tau}^\dagger \hat{c}_{i \tau'} = \frac{1}{2} (n_{i \uparrow} + n_{i \downarrow})\delta_{\tau \tau'}  + \bs{S}_i\cdot\bs{\tau}_{\tau', \tau}, \label{SpinOperatorInv1}\\
&\hat{c}_{i \tau} \hat{c}_{i \tau'}^\dagger = \frac{1}{2} (2 - n_{i \uparrow} - n_{i \downarrow}) \delta_{\tau \tau'} - \bs{S}_i\cdot\bs{\tau}_{\tau, \tau'}, \label{SpinOperatorInv2}
\end{align}
we find the spin Hamiltonian for 2D AFM Mott insulators,
\begin{align}
\label{MKMHeff0}
H_{\text{S}} =& J_{1}\sum_{\langle i,j \rangle} \bs{S}_i\cdot\bs{S}_j + J_{2}\sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i\cdot\bs{S}_j,  \n \\
&+ \sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i \bs{\Gamma}_{ij} \bs{S}_j +\sum_{\langle \langle i,j \rangle \rangle} \bs{D}_{ij}\cdot \bs{S}_i \times \bs{S}_j,
\end{align}
with the following spin-spin interactions,
\begin{subequations}
\label{spin-para}
\begin{align}
&J_{1(2)} = \frac{2t_{1(2)}^2}{U}, \\
&\bs{\Gamma}_{ij} =\frac{2\Delta^2}{U} \text{diag}(-1,-1,1),\\
&\bs{D}_{ij} = \frac{4 t_2 \Delta}{U}\nu_{ij}  \hat{\mathrm{e}}_z.
\end{align}
\end{subequations}
The first and second terms in the spin Hamiltonian, Eq. (\ref{MKMHeff0}), are the NN and NNN symmetric Heisenberg AFM exchange interactions, $J_{1(2)}>0$, respectively; the third term is the NNN anisotropic Heisenberg exchange interaction (XXZ-like term) arising from the intrinsic SOI, and finally, the last term is the intrinsic NNN DMI. The intrinsic SOI in Eq. (\ref{Hsoi}), leads to a DMI vector, $\bs{D}_{ij}$, perpendicular to the honeycomb plane. It can also be shown that the Rashba SOI results in an in-plane DMI vector that is perpendicular to the lattice bonds.
For the sake of completeness, we explain shortly the effect of disorder by adding an on-site disorder potential $\sum_{i \tau} \varepsilon_i \hat{c}_{i \tau}^\dagger \hat{c}_{i \tau}$ in the Kane-Mele-Hubbard Hamiltonian Eq. (\ref{MKMH}), where $\varepsilon_i$ is an uncorrelated random variable. Following the above procedure, it can be shown that the spin interaction parameters are renormalized as $U^{-1} \rightarrow U/\left(U^2-(\epsilon_j-\epsilon_i)^2\right)$ \cite{Protopopov2018}. In the large $U$ limit, the effect of disorder is negligible and we thus ignore it in the rest of this Letter.

The terms in the spin Hamiltonian derived in Eq. (\ref{MKMHeff0}) have already been obtained phenomenologically \cite{Owerre2016} and confirmed experimentally \cite{Chen2018}. To the best of our knowledge, this is the first microscopic derivation of the complete spin Hamiltonian $\hat{H}_{\text{S}}$ from the microscopic Kane-Mele-Hubbard model.
This Hamiltonian has interesting features and exotic phases like, the existence of magnon spin Nernst effect in collinear AFM layers \cite{Cheng2016, PhysRevLett.117.217203}, topological magnon insulator phase \cite{Owerre2016, Elyasi2018, Chen2018}, spin Hall effects of Weyl magnons \cite{Zyuzin2018, Sekine2016} magnonic Floquet topological insulators, spin density wave \cite{Mulder2010}, chiral and topological gapped spin liquid phases \cite{Vaezi}. Ultrafast control of DMI and exchange interaction by laser pulses enables engineering all these phenomena.

\textit{Laser illumination}.
The effect of laser irradiation is introduced in the Kane-Mele-Hubbard Hamiltonian, Eq. (\ref{MKMH}), via the Peierls substitution method \cite{Peierls1933}. The electric-field component of a polarized laser pulse is given by $\bs{E}(t) = \frac{E_0}{2}( e^{-i\omega t}\hat{\epsilon}+\mathrm{c.c.})$, where $E_0$ is the electric field amplitude, $\omega$ is the laser pulse frequency, and $\hat{\epsilon} = (\hat{\mathrm{e}}_x+i\lambda\hat{\mathrm{e}}_y)/\sqrt{1+\lambda^2}$ is the polarization unit vector with $\lambda= 0$ for linear polarization and $\lambda= \pm 1$ for right- and left handed polarizations.

It is more convenient to rewrite the noninteracting part of the Kane-Mele-Hubbard Hamiltonian, Eq. (\ref{MKMH}), as $\hat{T}_0=\hat{\mathcal{H}}_{\text{K}}+\hat{\mathcal{H}}_{\text{SOI}} = - \sum_{i,j , \tau, \tau'}
t_{ij}^{\tau\tau'} \hat{c}_{i \tau}^\dagger \hat{c}_{j \tau'}$, where the hopping amplitude is $t_{ij}^{\tau\tau'} = \delta_{\tau,\tau'}t_1$ for $i,j$ being NN and $t_{ij}^{\tau\tau'} = \delta_{\tau,\tau'}t_2 - i\Delta\nu_{ij}\sigma^z_{\tau, \tau'}$ for $i,j$ being NNN.
According to the Peierls substitution method, the hopping part of the Hamiltonian gain an extra phase $t_{ij}^{\tau\tau'}\rightarrow t_{ij}^{\tau\tau'} e^{{i e \bs{R}_{ij} \cdot \bs{A}(t)}}$, where $\bs{R}_{ij} = \bs{R}_i-\bs{R}_j$, $\bs{R}_i$ is the position of site $i$, $e$ is the electronic charge, and $\bs{A}$ is the vector potential of the laser pulse $\bs{A}(t) = \frac{1}{2}(\bs{A} e^{-i\omega t} + \mathrm{c.c.})$, with $\bs{A} = \frac{iE_0}{\omega}\hat{\epsilon}$.
The Peierls phase can be rewritten as $e\bs{R}_{ij}\cdot\bs{A} \equiv \alpha_{ij} e^{i \theta_{ij}}$, with $\alpha_{ij} = \pm|e\bs{R}_{ij}\cdot \bs{A}|$, in a such way that $\alpha_{ij}= -\alpha_{ji}$, $\theta_{ij}= \theta_{ji}$, and $\theta_{ij} \in \left[0,\pi\right)$.
Now, we use the Jacobi–-Anger expansion to rewrite the Peierls phase in the basis of its harmonics,
\begin{equation}
\label{JacobiAnger}
e^{ie\bs{R}_{ij}\cdot\bs{A}(t)} = \sum_m e^{i(\frac{\pi}{2}-\theta_{ij})m} \mathcal{J}_m(\alpha_{ij}) e^{im\omega t},
\end{equation}
where $\mathcal{J}_m(x)$ is the \textit{m}-th Bessel function of the first kind \cite{Kitamura2017}.

The total Hamiltonian in the presence of the laser field is time dependent through its hopping part, $\hat{H}(t) = \hat{T}(t) +  U\hat{d}$. Using Eq. (\ref{JacobiAnger}), we find $\hat{T}(t) = \sum_m \hat{T}_m e^{im \omega t}$, where $\hat{T}_m$ is the sum of all the \textit{m}-th Fourier mode of the hopping terms. We can additionally decompose $\hat{T}_m = \hat{T}_{-1,m}+\hat{T}_{0,m}+\hat{T}_{1,m}$, where $\hat{T}_{dm}(t)$ changes the doublon number by adding $d$ double occupancies, $\hat{T}_{dm}(t) = \sum_i \hat{P}_{i+d}\hat{T}_{m}(t)\hat{P}_i$. With this, we can write the hopping operator as
\begin{equation}
\hat{T}(t) = \sum_m (\hat{T}_{-1,m}+\hat{T}_{0,m}+\hat{T}_{1,m})e^{im\omega t},
\end{equation}

>>>>

% !TEX spellcheck = en-US
\documentclass[aps,prl,twocolumn,amsmath,amssymb,nobibnotes]{revtex4-1}%
\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{bm}
\usepackage{color}
\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage{braket}
%\usepackage{caption}
%\usepackage{subcaption}
\usepackage{fancybox}
\usepackage{amsfonts}
\usepackage{amssymb}%
\usepackage{natbib}
\usepackage[toc,page]{appendix}
\usepackage[symbol]{footmisc}
%\usepackage[style=numeric-comp]{biblatex}
\setcounter{MaxMatrixCols}{30}
%\usepackage{subcaption}

@@

\begin{document}
\title{Ultrafast Control of Spin Interactions in  Honeycomb Antiferromagnetic Insulators}

\author{Juan M. Losada}
\affiliation{Center for Quantum Spintronics, Department of Physics, Norwegian University of Science and Technology, NO-7491 Trondheim, Norway}
\author{Arne Brataas}
\affiliation{Center for Quantum Spintronics, Department of Physics, Norwegian University of Science and Technology, NO-7491 Trondheim, Norway}
\author{Alireza Qaiumzadeh}
\thanks{Corresponding author: alireza.qaiumzadeh@ntnu.no}
\affiliation{Center for Quantum Spintronics, Department of Physics, Norwegian University of Science and Technology, NO-7491 Trondheim, Norway}
%\affiliation{Institute for Advanced Studies in Basic Science (IASBS), Zanjan, Iran}

\begin{abstract}
Light enables ultrafast, direct and nonthermal control of the exchange and Dzyaloshinskii-Moriya interactions. We consider two-dimensional honeycomb lattices described by the Kane-Mele-Hubbard model at half filling and in the strongly correlated limit, i.e., the Mott insulator phase of a canted antiferromagnet. Based on Floquet theory, we demonstrate that by changing the amplitude and frequency of polarized laser pulses, one can tune the amplitudes and signs of and even the ratio between the exchange and Dzyaloshinskii-Moriya spin interactions. Furthermore, the renormalizations of the spin interactions are independent of helicity. Our results pave the way for ultrafast optical spin manipulation in recently discovered two-dimensional magnetic materials.
\end{abstract}

@@

\textit{Introduction}.--- The discovery of all-optical control of the order parameters in antiferromagnetic (AFM) and ferromagnetic (FM) materials by means of ultrashort intense high-frequency laser pulses has propelled spintronics into a new era of ultrafast magnetism \cite{kimel2005ultrafast,2009NatPh...5..727K,Rasing}.
Despite many attempts to uncover its origin, the detailed underlying microscopic mechanism remains unclear.
The process triggering ultrafast magnetization dynamics phenomena may rely on either thermal or nonthermal mechanisms \cite{heat-light,Rasing}. Thus far, it has been believed that ultrafast {\it{nonthermal}} manipulation of spins is possible only through either a direct coupling between the magnetic field component of the laser pulses and the spins or an indirect coupling between the electric field component of the light and the spins via spin-orbit coupling \cite{Rasing,Mikhaylovskiy2015,AlirezaQ1,AlirezaQ2}. Recent experiments have demonstrated that laser pulses can directly modify the amplitude and sign of the exchange interaction, which is the strongest spin interaction in magnetically ordered systems \cite{2018Natur.553..481G,Mikhaylovskiy2015,Perakis,PhysRevLett.93.197403}.
Refs. \onlinecite{PhysRevLett.113.057201,PhysRevLett.115.075301,Mentink2015,PhysRevLett.118.097701,Stepanov2017,Kitamura,MentinkReview,Barbeau2018} have theoretically proposed that a direct coupling between the electric field of the light and the spins facilitates nonthermal optical modification of the exchange interaction, in agreement with recent experiments \cite{2018Natur.553..481G,Mikhaylovskiy2015}.

In magnetic systems with broken inversion symmetry, there is also an antisymmetric exchange interaction between spins that breaks the chiral symmetry, namely, the Dzyaloshinskii-Moriya interaction (DMI) \cite{dzyaloshinsky1958, moriya1960,Alireza-DMI1, Alireza-DMI2}. Although this interaction is considerably weaker than the exchange interaction, it is essential in magnetic materials for enabling weak ferromagnetism in AFM materials \cite{dzyaloshinsky1958, moriya1960}, topological objects such as chiral skyrmions \cite{2009Sci...323..915M,2017NatRM...217031F,AlirezaSk1, AlirezaSk2} and chiral domain walls \cite{Thiaville_2012,Parkin,AlirezaSW}, and exotic phases of topological magnon insulators \cite{Owerre2016,PhysRevLett.117.227201, Elyasi2018, Chen2018, Owerre1, Owerre2}. The ratio between the exchange interaction and the DMI controls the tilt angle of the canted spins. Finding a mechanism for tuning this ratio can enable new phenomena in ultrafast spin dynamics and switching \cite{Mikhaylovskiy2015,Mentink2015,Stepanov2017,Kitamura}.

Another far-reaching recent breakthrough in spintronics is the discovery of two-dimensional (2D) Van der Waals AFM and FM materials with metallic, semiconducting and insulating band structures \cite{2D1,2D2}. The advantages presented by the existence of low dimensionality and magnetic order in the same material enable the development of new spintronics devices with exceptional performance.

In this Letter, we show that intense high-frequency laser pulses can dramatically affect the spin-spin interactions and the ratio between the exchange interaction and the DMI in a broad class of 2D magnetic materials described by the Kane-Mele-Hubbard model.  We find that light can be used to tune both the sign and magnitude of the AFM exchange interaction, in agreement with the dynamical mean-field theory \cite{Mentink2015}. Importantly, we demonstrate that laser pulses can also be used to independently change the sign and magnitude of the DMI, thus enabling rapid control of magnetism. The ability to independently control the signs and magnitudes of the exchange interaction and the DMI enables superior control of magnetic textures in 2D magnets.

\textit{Model Hamiltonian}.---
The electron dynamics in 2D planar honeycomb lattices can be described by the Kane-Mele-Hubbard model \cite{Kane2005,PhysRevLett.107.010401,PhysRevLett.106.100403,PhysRevB.82.075106,PhysRevB.84.235149,Griset2012,Auerbach}; see Fig. \ref{fig1}. In the absence of external perturbations, the Hamiltonian is the sum of the kinetic term $\hat{\mathcal{H}}_t$, the intrinsic spin-orbit interaction (SOI) $\hat{\mathcal{H}}_{\text{SOI}}$, and the repulsive Coulomb interaction between the electrons as modeled in the form of the extended Hubbard interaction $\hat{\mathcal{H}}_{\text{int}}$:
\begin{equation}
\label{MKMH}
\hat{H}_0 = \hat{\mathcal{H}}_{\text{K}} + \hat{\mathcal{H}}_{\text{SOI}} + \hat{\mathcal{H}}_{\text{int}},
\end{equation}
where
\begin{align}
&\hat{\mathcal{H}}_{\text{K}} = - t_{1}\sum_{\langle i,j \rangle, \tau} \hat{c}_{i \tau}^\dagger \hat{c}_{j \tau} - t_{2}\sum_{\langle \langle i,j \rangle \rangle, \tau} \hat{c}_{i \tau}^\dagger \hat{c}_{j \tau}, \label{Ht} \\
&\hat{\mathcal{H}}_{\text{SOI}}  = i \Delta \sum_{\langle \langle i,j \rangle \rangle, \tau,\tau'} \nu_{ij}\sigma^z_{\tau, \tau'}\hat{c}_{i \tau}^\dagger \hat{c}_{j \tau'} \label{Hsoi}, \\
&\hat{\mathcal{H}}_{\text{int}} = U_{00}\sum_{i=1} \hat{n}_{i\uparrow}\hat{n}_{i\downarrow} + \frac{1}{2}\sum_{\langle i,j \rangle, \tau \tau'} V_{ij}\hat{n}_{i\tau}\hat{n}_{j\tau'} \label{Hint}.
\end{align}
Here, $\langle\cdot\rangle$ and $\langle\langle\cdot\rangle\rangle$ denote nearest neighbors (NN) and next-nearest neighbors (NNN), respectively; $\hat{c}_{i \tau}^\dagger$ and $ \hat{c}_{i \tau}$ are the fermionic creation and annihilation operators, respectively, for an electron at site $i$ and in spin state $\tau=\{\uparrow,\downarrow\}$;
$t_1$ and $t_2$ are the NN and NNN hopping amplitudes, respectively; $\Delta$ is the intrinsic SOI parameter, and $\nu_{ij}=\pm 1$ depending the hoping orientation from $j$ to $i$ (see Fig. \ref{fig1}); $\sigma^{z}$ is the $z$-component of the Pauli matrices $\bm{\sigma}$; and $U_{00}$ and $V_{ij}$ are the on-site and NN Coulomb interactions, respectively.
The intrinsic NNN SOI, Eq. (\ref{Hsoi}), reduces the $\mathrm{SU}(2)$ symmetry of the original Hubbard model to the $\mathrm{U}(1)$ spin group. In buckled non-coplanar honeycomb lattices or systems with structural inversion asymmetry, the presence of NNN or NN Rashba SOIs, respectively, further reduces the symmetry to $Z_2$.

\begin{figure}[t]
\centering
\includegraphics[width=0.6\columnwidth]{kmh.png}
\caption{A honeycomb cell with NN hopping $t_1$, NNN hopping $t_2$, intrinsic SOI $\Delta$, and $\nu_{ij} = \pm 1$ for clockwise and counterclockwise hopping.}
\label{fig1}
\vspace*{-6pt}
\end{figure}

Using the variational principle, it has been shown that the NN Coulomb interaction can be approximated by a renormalized local interaction $U =U_{00} - \bar{V}$, where $\bar{V}$ is a weighted average of the NN Coulomb interaction \cite{Schuler2013}. Thus, we consider only the local Coulomb interaction in our total Hamiltonian and express the interaction part of the Hamiltonian in Eq. (\ref{Hint}) as
\begin{equation}
\hat{\mathcal{H}}_{\text{int}} \approx U\hat{d},
\end{equation}
where we introduce the doublon number operator $\hat{d} = \sum_{i=1} \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}$, which has an eigenvalue of $d$.
For later use, we define $\hat{P}_d$ as the projection operator onto the subspace spanned by states with eigenvalue $d$, i.e., states with exactly $d$ doublons. At site $i$, we can define the projection operator related to double occupancy as $\hat{P}_{i,1} = \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}$ and the projection operator related to the absence of double occupancy as $\hat{P}_{i,0} = 1 - \hat{P}_{i,1}$ \cite{Fazekas}. We can then define the projection operator $\hat{P}_{d}$ for the whole system as follows. Let $\mathcal{O}$ and $\mathcal{P}^d(\mathcal{O})$ denote the set of sites on the lattice and the set of subsets of $\mathcal{O}$ with exactly $d$ elements, respectively. Then, in compact form, the projection operator reads $\hat{P}_d = \sum_{A \in \mathcal{P}^d(\mathcal{O})} \left\{\prod_{i \in A} \hat{P}_{i,1} \prod_{i \notin A} \hat{P}_{i,0} \right\}$.

We are interested in the strongly correlated regime $U\gg t_{1(2)}$ at half filling. In the limit of such strong coupling, any state with a nonzero number of double occupancies ($d \neq 0$) has a much larger energy than a state with no double occupancy ($d=0$). We obtain an effective Hamiltonian acting on the $d=0$ subspace by means of second-order perturbation theory in the hopping terms. Using the relations
\begin{subequations}
\label{SpinOperatorInv1}
\begin{align}
&\hat{c}_{i \tau}^\dagger \hat{c}_{i \tau'} = \frac{1}{2} (n_{i \uparrow} + n_{i \downarrow})\delta_{\tau \tau'}  + \bs{S}_i\cdot\bs{\tau}_{\tau', \tau}, \\
&\hat{c}_{i \tau} \hat{c}_{i \tau'}^\dagger = \frac{1}{2} (2 - n_{i \uparrow} - n_{i \downarrow}) \delta_{\tau \tau'} - \bs{S}_i\cdot\bs{\tau}_{\tau, \tau'},
\end{align}
\end{subequations}
we find the spin Hamiltonian for 2D AFM Mott insulators,
\begin{align}
\label{MKMHeff0}
H_{\text{S}} =& J_{1}\sum_{\langle i,j \rangle} \bs{S}_i\cdot\bs{S}_j + J_{2}\sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i\cdot\bs{S}_j,  \n \\
&+ \sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i \bs{\Gamma} \bs{S}_j +\sum_{\langle \langle i,j \rangle \rangle} \bs{D}_{ij}\cdot \bs{S}_i \times \bs{S}_j,
\end{align}
with the following spin-spin interactions:
\begin{subequations}
\label{spin-para}
\begin{align}
&J_{1(2)} = \frac{2t_{1(2)}^2}{U}, \\
&\bs{\Gamma} =\frac{2\Delta^2}{U} \text{diag}(-1,-1,1),\\
&\bs{D}_{ij} = \frac{4 t_2 \Delta}{U}\nu_{ij}  \hat{\mathrm{e}}_z.
\end{align}
\end{subequations}
In the spin Hamiltonian given in Eq. (\ref{MKMHeff0}), the first and second terms are the NN and NNN symmetric Heisenberg AFM exchange interactions ($J_{1(2)}>0$), respectively; the third term is the NNN anisotropic Heisenberg exchange interaction (an XXZ-like term) arising from the intrinsic SOI; and the last term is the intrinsic NNN DMI. The intrinsic SOI in the Kane-Mele-Hubbard model, Eq. (\ref{Hsoi}), leads to a DMI vector, $\bs{D}_{ij}$, that is perpendicular to the honeycomb plane. It can also be shown that the Rashba SOI results in an in-plane DMI vector that is perpendicular to the lattice bonds.

Although the microscopic derivation of anisotropic exchange interaction in the spin Hamiltonian of Eq. (\ref{MKMHeff0}) has been reported before \cite{PhysRevB.82.075106,Vaezi,2017TDM.....4c5002L}, we are not aware of any other microscopic calculation of intrinsic NNN DMI in honeycomb lattices \cite{Owerre2016,PhysRevLett.117.227201,Chen2018}. The spin Hamiltonian of Eq. (\ref{MKMHeff0}) gives rise to several interesting features and exotic phases, such as the existence of the magnon spin Nernst effect in collinear AFM layers \cite{Cheng2016, PhysRevLett.117.217203}, a topological magnon insulator phase \cite{Owerre2016, PhysRevLett.117.227201, Elyasi2018, Chen2018}, spin Hall effects for Weyl magnons \cite{Zyuzin2018, Sekine2016}, magnonic Floquet topological insulators, spin density waves \cite{Mulder2010}, and chiral and topological gapped spin liquid phases \cite{Vaezi}. Ultrafast control of the DMI and the exchange interaction by means of laser pulses can enable engineering of all of these phenomena.

For completeness, let us briefly illustrate the effect of disorder by adding an on-site disorder potential $\sum_{i \tau} \varepsilon_i \hat{c}_{i \tau}^\dagger \hat{c}_{i \tau}$ to the Kane-Mele-Hubbard Hamiltonian given in Eq. (\ref{MKMH}), where $\varepsilon_i$ is an uncorrelated random variable. Following the above procedure, it can be shown that the spin interaction parameters are renormalized as $1/U \rightarrow 1/\left(U-(\epsilon_j-\epsilon_i)^2/U\right)$ \cite{Protopopov2018}. In the large $U$ limit and in the presence of very high frequency oscillations, the effect of disorder is negligible; thus, we do not include it in the following.

\textit{Laser illumination}.---
We introduce the effects of laser irradiation in the Kane-Mele-Hubbard Hamiltonian of Eq. (\ref{MKMH}) via the Peierls substitution \cite{Peierls1933}. The electric field of a polarized laser pulse is  $\bs{E}(t) = E_0( e^{-i\omega t}\hat{\epsilon}+\mathrm{c.c.})/2$, where $E_0$ is the electric field amplitude, $\omega$ is the laser pulse frequency, and $\hat{\epsilon} = (\hat{\mathrm{e}}_x+i\lambda\hat{\mathrm{e}}_y)/\sqrt{1+\lambda^2}$ is the unit vector representing the laser polarization, with $\lambda= 0$ for linear polarization, and $\lambda= \pm 1$ for right- and left-handed polarizations.
